\pkg{casebase}: An Alternative Framework For Survival Analysis

### 2019-12-11 analysis/paper/paper.Rmd
edit third case study

---
documentclass: jss
author:
  - name: Sahir Bhatnagar *
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{sahir.bhatnagar@mail.mcgill.ca}
    url: http://sahirbhatnagar.com/
  - name: Maxime Turgeon *
    affiliation: 'University of Manitoba \AND'
    address: >
      186 Dysart Road
      Winnipeg, MB, Canada R3T 2N2
    email: \email{max.turgeon@umanitoba.ca}
    url: https://maxturgeon.ca/
  - name: Jesse Islam 
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{jesse.islam@mail.mcgill.ca}
  - name: James Hanley
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{james.hanley@mcgill.ca}
    url: http://www.medicine.mcgill.ca/epidemiology/hanley/
  - name: Olli Saarela
    affiliation: University of Toronto
    address: >
      Dalla Lana School of Public Health, 
      155 College Street, 6th floor, 
      Toronto, Ontario 
      M5T 3M7, Canada
    email: \email{olli.saarela@utoronto.ca}
    url: http://individual.utoronto.ca/osaarela/
title:
  formatted: "\\pkg{casebase}: An Alternative Framework For Survival Analysis"
  # If you use tex in the formatted title, also supply version without
  plain:     "casebase: An Alternative Framework For Survival Analysis"
  # For running headers, if needed
  short:     "\\pkg{casebase}: An Alternative Framework For Survival Analysis"
abstract: >
  The abstract of the article. * joint co-authors
keywords:
  # at least one keyword must be supplied
  formatted: [keywords, not capitalized, "\\proglang{Java}"]
  plain:     [keywords, not capitalized, Java]
preamble: >
  \usepackage{amsmath}
  \usepackage{longtable}
  \usepackage{graphicx}
output: 
  rticles::jss_article:
    number_sections: TRUE     #added argument option 
    citation_package: "natbib"  #All my citations use biblatex, not natbib. 
biblio-style: jss      #Listed to use in JSS Instructions for Authors, but not in template by default. 
bibliography: references.bib  #Also not included in template by default. 
fig_width: 7
fig_height: 6
fig_caption: true
---

@@

```{r echo=FALSE, eval = FALSE}
table1 <- tibble::tribble(
    ~Variable, ~Description, ~`Statistical summary`,
    "Sex", "Sex", "M=Male (100)",
    "", "", "F=Female (77)",
    "D", "Disease", "ALL (73)",
    "", "", "AML (104)",
    "Phase", "Phase", "CR1 (47)", 
    "", "", "CR2 (45)", 
    "", "", "CR3 (12)", 
    "", "", "Relapse (73)",
    "Source", "Type of transplant", "BM+PB (21)", 
    "", "", "PB (156)",
    "Age", "Age of patient (years)", "4–62", 
    "", "", "30.47 (13.04)",
    "Ftime", "Failure time (months)", "0.13–131.77",
    "", "", "20.28 (30.78)",
    "Status", "Status indicator", "0=censored (46)", 
    "", "", "1=relapse (56)", 
    "", "", "2=competing event (75)"
)

@@

<!-- latex table generated in R 3.4.1 by xtable 1.8-2 package -->
<!-- Sun Sep 17 16:03:22 2017 -->
\begin{table}[ht]
\centering
\begin{tabular}{ccc}
  \hline
Variable & Description & Statistical summary \\ 
  \hline
Sex & Sex & M=Male (100) \\ 
   &  & F=Female (77) \\ 
  D & Disease & ALL (73) \\ 
   &  & AML (104) \\ 
  Phase & Phase & CR1 (47) \\ 
   &  & CR2 (45) \\ 
   &  & CR3 (12) \\ 
   &  & Relapse (73) \\ 
  Source & Type of transplant & BM+PB (21) \\ 
   &  & PB (156) \\ 
  Age & Age of patient (years) & 4–62 \\ 
   &  & 30.47 (13.04) \\ 
  Ftime & Failure time (months) & 0.13–131.77 \\ 
   &  & 20.28 (30.78) \\ 
  Status & Status indicator & 0=censored (46) \\ 
   &  & 1=relapse (56) \\ 
   &  & 2=competing event (75) \\ 
   \hline
\end{tabular}
\caption{Baseline characteristics of patients in the stem-cell transplant study.}
\label{tab:table1bmtcrr}
\end{table}

@@

# Case study 3--Study to Understand Prognoses Preferences Outcomes and Risks of Treatment

We examined the Study to Understand Prognoses Preferences Outcomes and Risks of Treatment (SUPPORT) dataset (REF from main site). the SUPPORT dataset tracks death for individuals who are considered seriously ill within a hospital.  Imputation was conducted using the default methods in the mice package in R, available on CRAN. Before imputation, there was a total of 1000 individuals and 35 variables. After imputation, we have 474 observations where 67.5% of individuals in the study experience the event of interest. The SUPPORT dataset will be used to demonstrate regularization within casebase (REF), while comparing their absolute risk predictions for a new individual. A description of each variable can be found in Table~\ref{tab:support1} and a breakdown of each categorical variable in Table~\ref{tab:support2}. The data was downloaded from the main author's website, and it is also available as part of the casebase package.

@@

\caption{A description of each variable in the SUPPORT dataset, taken from (REF).}

@@

\begin{table}[ht]
\centering
\begin{tabular}{cc}
\hline
\textbf{Variable} & \textbf{Levels}                      \\
\hline
sex      & female                                        \\
         & male                                          \\
dzgroup  & ARF/MOSF w/Sepsis                             \\
         & COPD                                          \\
         & CHF                                           \\
         & Cirrhosis                                     \\
         & Coma                                          \\
         & Colon Cancer                                  \\
         & Lung Cancer                                   \\
         & MOSF w/Malig                                  \\
dzclass  & ARF/MOSF                                      \\
         & COPD/CHF/Cirrhosis                            \\
         & Coma                                          \\
         & Cancer                                        \\
income   & under \$11k                                   \\
         & $11-$25k                                      \\
         & $25-$50k                                      \\
         & \textgreater{}\$50k                           \\
race     & white                                         \\
         & black                                         \\
         & asian                                         \\
         & other                                         \\
         & hispanic                                      \\
sfdm2    & no(M2 and SIP pres)                           \\
         & adl\textgreater{}=4 (\textgreater{}=5 if sur) \\
         & SIP\textgreater{}=30                          \\
         & Coma or Intub                                 \\
         & \textless{}2 mo. follow-up                   
\end{tabular}
\caption{A description of each level within each categorical variable in the dataset, taken from REF.}
\label{tab:support2}
\end{table}

Hospital death as a covariate was removed, as this is directly informative of death. The glmnet package(REF) on CRAN was used to introduce elastic net, lasso and ridge regressions to casebase. To visualize the effect of imputation on incidence density, we can refer to the pre-imputation population-time plot in Figure \ref{fig:ptPreI} and post-imputation population time plot in Figure \ref{fig:ptPostI}. The incidence density is consistent between imputed and complete datasets.

```{r supportSetUp, echo = FALSE, eval = eval_cs3}
completeData=casebase::support
#keep one individual out, at random, for which we will develop an 
#absolute risk curve.
sam=sample(1:nrow(completeData), 1)
#Set workingData to the remaining individuals in the data
#workingData=completeData[-c(sam),]
workingCompleteData=as.matrix(sparse.model.matrix(death~ .-d.time,data=completeData))[,-c(1)]
#Create u and xCox, which will be used when fitting Cox with glmnet.
newData=workingCompleteData[sam,]
x=workingCompleteData[-c(sam),]
#x and y will be used to fit the casebase model under glmnet.
#x=as.matrix(sparse.model.matrix(death~ .-d.time+0,data=workingData))
y=data.matrix(completeData[-c(sam),c(34,4)])
```

```{r poptimePreImputation, echo = TRUE, eval = eval_cs3, fig.cap="\\label{fig:ptPreI} Population-time plot before imputation. Red dots demonstrate death, while each gray dot makes up all the person moments spent in the study."}
#demonstrating incidence density before imputation
plot(casebase::popTime(as.data.frame(support),time = "d.time",event = "death"))
```

```{r poptimePostImputation, echo = TRUE, eval = eval_cs3, fig.cap="\\label{fig:ptPostI} Population-time plot before imputation. Red dots demonstrate death, while each gray dot makes up all the person moments spent in the study."}
#demonstrating incidence density after imputation
plot(casebase::popTime(as.data.frame(workingCompleteData),time = "d.time",event = "death"))
```
Our main objective is to compute the absolute risk of death for a given set of covariates, using regularization when fitting our model. First, we fit a smooth hazard to the data; for the sake of this example, we opted for a linear term for time. As casebase makes use of the glmnet package, we interact with the fitsmoothhazard.fit function using a matrix interface, where y contains the time and event variables and x contains all other variables we would like to include in the model. For the SUPPORT dataset, all factor variables were converted into dummy variables. Here, we used lasso penalization by setting alpha to 1.
```{r CBhazard, echo = TRUE, eval = eval_cs3}
#Using casebase's glmnet implementation, using splines and a case to base ratio of 100,
#to determine the hazard function
cb.Model=casebase::fitSmoothHazard.fit(x,y,family="glmnet",time="d.time",event="death",
                             formula_time = ~d.time,alpha=1,ratio=100)
```
Then, we take our model and use the absoluteRisk function to integrate and retrieve our absolute risk function.
```{r CBabsolute, echo = TRUE, eval = eval_cs3}
#Estimating the absolute risk curve using the newData parameter.
casebaseAbsolute=casebase::absoluteRisk(cb.Model,time = seq(1,max(completeData$d.time), 0.5),newdata = t(newData) ,
                              s="lambda.1se",method=c("numerical"))
```
We will compare this curve to a regularized version of cox regression, and a kaplan-meier survival curve. The regularized cox regression absolute risk curve required two extra steps to retrieve. Coxnet from the glmnet package has tools to fit a regularized cox model, but requires that a survival object skeleton of the selected variables, so that the survival package tools can be used to retrieve the absolute risk. We handle this by fitting a regularized cox model with the glmnet package, and a cox model from the survival package with the selected variables from the regularized fit. This second model serves as a skeleton that permits the use of survival package functions. The coefficients from the regularized model will replace the coefficients in our skeleton. the resulting model will have the correct coefficients when integrating, but will have an incorrect standard error. For the purposes of this case study, we are only interested in the absolute risk curve itself and not the standard error.

```{r coxHazAbsolute, echo = TRUE, eval = eval_cs3}
#Create u and xCox, which will be used when fitting Cox with glmnet.
u=survival::Surv(time = as.numeric(y[,2]), event = as.factor(y[,1]))
xCox=as.matrix(sparse.model.matrix(death~ .-d.time,data=completeData))[-c(sam),]
#hazard for cox using glmnet
coxglmFit=glmnet::cv.glmnet(x=xCox,y=u, family="cox",alpha=1)
#convergence demonstrated in plot
plot(coxglmFit)
#taking the coefficient estimates for later use
nonzero_covariate_cox <- predict(coxglmFit, type = "nonzero", s = "lambda.1se")
nonzero_coef_cox <- coef(coxglmFit, s = "lambda.1se")
#creating a new dataset that only contains the covariates chosen through glmnet.
#cleanCoxData<- as.data.frame(cbind(as.numeric(workingData$d.time),as.factor(workingData$death), xCox[,nonzero_covariate_cox$X1]))
cleanCoxData<-as.data.frame(cbind(as.numeric(y[,2]),as.numeric(y[,1]),xCox[,nonzero_covariate_cox$X1]))
#newDataCox<-xCox[sam,nonzero_covariate_cox$X1]
#fitting a cox model using regular estimation, however we will not keep it.
#this is used more as an object place holder.
coxFit <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = cleanCoxData)
#The coefficients of this object will be replaced with the estimates from coxglmFit.
#Doing so makes it so that everything is invalid aside from the coefficients.
#In this case, all we need to estimate the absolute risk is the coefficients.
#Std. error would be incorrect here, if we were to draw error bars.
coxFit$coefficients<-nonzero_coef_cox@x
#Fitting absolute risk curve for cox+glmnet
newDataCox=newData[nonzero_covariate_cox$X1-1]

abCoxFit<-survival::survfit(coxFit,newdata=as.data.frame(t(newDataCox)),time = seq(0,max(completeData$d.time), 1),type="breslow")
```
We used the survival package to calculate the kaplan-meier absolute risk function. 
```{r fitKM+abRisk, echo = TRUE, eval = eval_cs3}
#creating a surv object to be used to fit an unadjusted absolute risk curve.
# (Kaplan-Meier risk curve)
km <- survival::Surv(time = completeData$d.time, event = completeData$death)
abKm<-survival::survfit(km~1,type='kaplan-meier',conf.type='log')
```

Now that our three absolute risk functions have been calculated, we compare them all in Figure \ref{fig:abSC}. Both coxnet and casebase decrease the absolute risk by the end of the study, in comparison to kaplan-meier.
```{r abSupportComparison, echo = TRUE, eval = eval_cs3, fig.cap="\\label{fig:abSC} Compares regularized casebase in black, regularized cox in red and Kaplan meier in blue."}
# A plot to compare all three Absolute risk (Commulative Incidence)
plot(casebaseAbsolute,type='l',col="black",lwd=2,main="Support- CaseBase vs. Cox+glmnet vs KM Absolute Risk Curves",xlab="Survival-Time",ylab="Cumulative Incidence",ylim=c(0,1),xlim=)
lines(abCoxFit ,col="red",fun="event",lwd=3,conf.int = FALSE)
lines(abKm ,col="Blue",fun="event",lwd=3,conf.int = FALSE)
legend("bottom", 
       legend = c( "Casebase (Lasso+linear)","semi-parametric (Cox)+glmnet","KM curve"), 
       col = c("black","red","Blue"),
       lty = c(1, 1, 1), 
       bg = "gray90")
```

@@

The purpose of the next case study is to show how \pkg{casebase} can also be used for survival analysis problems that do not share the properties above. To this end, we will use the Stanford Heart Transplant Data [@clark1971cardiac,@crowley1977covariance]; we will use the version available in the \pkg{survival} package.

@@

In the following table we provide a comparison between the Cox model and case-base sampling:

>>>>

---
documentclass: jss
author:
  - name: Sahir Bhatnagar *
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{sahir.bhatnagar@mail.mcgill.ca}
    url: http://sahirbhatnagar.com/
  - name: Maxime Turgeon *
    affiliation: 'University of Manitoba \AND'
    address: >
      186 Dysart Road
      Winnipeg, MB, Canada R3T 2N2
    email: \email{max.turgeon@umanitoba.ca}
    url: https://maxturgeon.ca/
  - name: Jesse Islam 
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{jesse.islam@mail.mcgill.ca}
  - name: James Hanley
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{james.hanley@mcgill.ca}
    url: http://www.medicine.mcgill.ca/epidemiology/hanley/
  - name: Olli Saarela
    affiliation: University of Toronto
    address: >
      Dalla Lana School of Public Health, 
      155 College Street, 6th floor, 
      Toronto, Ontario 
      M5T 3M7, Canada
    email: \email{olli.saarela@utoronto.ca}
    url: http://individual.utoronto.ca/osaarela/
title:
  formatted: "\\pkg{casebase}: An Alternative Framework For Survival Analysis"
  # If you use tex in the formatted title, also supply version without
  plain:     "casebase: An Alternative Framework For Survival Analysis"
  # For running headers, if needed
  short:     "\\pkg{casebase}: An Alternative Framework For Survival Analysis"
abstract: >
  The abstract of the article. * joint co-authors
keywords:
  # at least one keyword must be supplied
  formatted: [keywords, not capitalized, "\\proglang{Java}"]
  plain:     [keywords, not capitalized, Java]
preamble: >
  \usepackage{amsmath}
  \usepackage{longtable}
  \usepackage{graphicx}
  \DeclareUnicodeCharacter{2500}{-}
output: 
  rticles::jss_article:
    number_sections: TRUE     #added argument option 
    citation_package: "natbib"  #All my citations use biblatex, not natbib. 
biblio-style: jss      #Listed to use in JSS Instructions for Authors, but not in template by default. 
bibliography: references.bib  #Also not included in template by default. 
fig_width: 7
fig_height: 6
fig_caption: true
---

@@

```{r echo=FALSE, eval = FALSE}
table1 <- tibble::tribble(
    ~Variable, ~Description, ~`Statistical summary`,
    "Sex", "Sex", "M=Male (100)",
    "", "", "F=Female (77)",
    "D", "Disease", "ALL (73)",
    "", "", "AML (104)",
    "Phase", "Phase", "CR1 (47)", 
    "", "", "CR2 (45)", 
    "", "", "CR3 (12)", 
    "", "", "Relapse (73)",
    "Source", "Type of transplant", "BM+PB (21)", 
    "", "", "PB (156)",
    "Age", "Age of patient (years)", "4--62", 
    "", "", "30.47 (13.04)",
    "Ftime", "Failure time (months)", "0.13--131.77",
    "", "", "20.28 (30.78)",
    "Status", "Status indicator", "0=censored (46)", 
    "", "", "1=relapse (56)", 
    "", "", "2=competing event (75)"
)

@@

<!-- latex table generated in R 3.4.1 by xtable 1.8-2 package -->
<!-- Sun Sep 17 16:03:22 2017 -->
\begin{table}[ht]
\centering
\begin{tabular}{ccc}
  \hline
Variable & Description & Statistical summary \\ 
  \hline
Sex & Sex & M=Male (100) \\ 
   &  & F=Female (77) \\ 
  D & Disease & ALL (73) \\ 
   &  & AML (104) \\ 
  Phase & Phase & CR1 (47) \\ 
   &  & CR2 (45) \\ 
   &  & CR3 (12) \\ 
   &  & Relapse (73) \\ 
  Source & Type of transplant & BM+PB (21) \\ 
   &  & PB (156) \\ 
  Age & Age of patient (years) & 4--62 \\ 
   &  & 30.47 (13.04) \\ 
  Ftime & Failure time (months) & 0.13--131.77 \\ 
   &  & 20.28 (30.78) \\ 
  Status & Status indicator & 0=censored (46) \\ 
   &  & 1=relapse (56) \\ 
   &  & 2=competing event (75) \\ 
   \hline
\end{tabular}
\caption{Baseline characteristics of patients in the stem-cell transplant study.}
\label{tab:table1bmtcrr}
\end{table}

@@

# Case study 3--SUPPORT Data

We examined the Study to Understand Prognoses Preferences Outcomes and Risks of Treatment (SUPPORT) dataset [-@knaus1995support]. The SUPPORT dataset tracks death for individuals who are considered seriously ill within a hospital. Imputation was conducted using the default methods in the \pkg{mice} package in \proglang{R}, available on CRAN. \textcolor{red}{Why was this method used? The Vanderbilt University website lists describes another method for imputation that we could have used.} Before imputation, there was a total of 1000 individuals and 35 variables. After imputation, we have 474 observations where 67.5% of individuals in the study experience the event of interest. The SUPPORT dataset will be used to demonstrate regularization within casebase, while comparing their absolute risk predictions for a new individual. A description of each variable can be found in Table~\ref{tab:support1} and a breakdown of each categorical variable in Table~\ref{tab:support2}. The data was downloaded from the main author's website, and it is also available as part of the \pkg{casebase} package.

@@

\caption{A description of each variable in the SUPPORT dataset.}

@@

\begin{table}[ht]
\centering
\begin{tabular}{cc}
\hline
\textbf{Variable} & \textbf{Levels}                      \\
\hline
sex      & female                                        \\
         & male                                          \\
dzgroup  & ARF/MOSF w/Sepsis                             \\
         & COPD                                          \\
         & CHF                                           \\
         & Cirrhosis                                     \\
         & Coma                                          \\
         & Colon Cancer                                  \\
         & Lung Cancer                                   \\
         & MOSF w/Malig                                  \\
dzclass  & ARF/MOSF                                      \\
         & COPD/CHF/Cirrhosis                            \\
         & Coma                                          \\
         & Cancer                                        \\
income   & under \$11k                                   \\
         & $11-$25k                                      \\
         & $25-$50k                                      \\
         & \textgreater{}\$50k                           \\
race     & white                                         \\
         & black                                         \\
         & asian                                         \\
         & other                                         \\
         & hispanic                                      \\
sfdm2    & no(M2 and SIP pres)                           \\
         & adl\textgreater{}=4 (\textgreater{}=5 if sur) \\
         & SIP\textgreater{}=30                          \\
         & Coma or Intub                                 \\
         & \textless{}2 mo. follow-up                   
\end{tabular}
\caption{A description of each level within each categorical variable in the dataset.}
\label{tab:support2}
\end{table}

Hospital death as a covariate was removed, as this is directly informative of death. The \pkg{glmnet} package on CRAN was used to introduce elastic net, lasso and ridge regressions to \pkg{casebase}. To visualize the effect of imputation on incidence density, we can refer to the pre-imputation population-time plot in Figure \ref{fig:ptPreI} and post-imputation population time plot in Figure \ref{fig:ptPostI}. The incidence density is consistent between imputed and complete datasets.

```{r supportSetUp, echo = FALSE, eval = eval_cs3}
completeData <- casebase::support
#keep one individual out, at random, for which we will develop an 
#absolute risk curve.
sam <- sample(1:nrow(completeData), 1)
#Set workingData to the remaining individuals in the data
#workingData=completeData[-c(sam),]
workingCompleteData <- as.matrix(sparse.model.matrix(death~ . - d.time, data = completeData))[,-c(1)]
#Create u and xCox, which will be used when fitting Cox with glmnet.
newData <- workingCompleteData[sam,]
x <- workingCompleteData[-c(sam),]
#x and y will be used to fit the casebase model under glmnet.
#x=as.matrix(sparse.model.matrix(death~ .-d.time+0,data=workingData))
y <- data.matrix(completeData[-c(sam), c(34, 4)])
```

```{r poptimePreImputation, echo = TRUE, eval = eval_cs3, fig.cap="\\label{fig:ptPreI} Population-time plot before imputation. Red dots demonstrate death, while each gray dot makes up all the person moments spent in the study."}
popTime_pre <- popTime(as.data.frame(support),
                       time = "d.time", event = "death")
plot(popTime_pre)
```

```{r poptimePostImputation, echo = TRUE, eval = eval_cs3, fig.cap="\\label{fig:ptPostI} Population-time plot before imputation. Red dots demonstrate death, while each gray dot makes up all the person moments spent in the study."}
popTime_post <- popTime(as.data.frame(workingCompleteData),
                        time = "d.time", event = "death")
plot(popTime_post)
```

Our main objective is to estimate the cumulative incidence of death for a given covariate profile. Given th number of covariates available, we will do this using regularized estimation. First, we fit the hazard function; for the sake of simplicity, we opted for a linear term for time. As \pkg{casebase} makes use of the \pkg{glmnet} package, we will use the function \code{fitsmoothhazard.fit}, which follows the syntax from \code{glmnet}. Specifically, we input a matrix \code{y}, which contains the time and event variables, and a matrix \code{x}, which contains all other variables we would like to include in the model. For the SUPPORT dataset, all factor variables were converted into dummy variables. Here, we used lasso penalization by setting \code{alpha} to 1.

```{r CBhazard, echo = TRUE, eval = eval_cs3}
cb.Model <- fitSmoothHazard.fit(x, y, family = "glmnet", time = "d.time", 
                                event = "death", formula_time = ~ d.time, 
                                alpha = 1, ratio = 100)
```

Then, we take our model object and use the \code{absoluteRisk} function to compute our estimate of the cumulative incidence function.

```{r CBabsolute, echo = TRUE, eval = eval_cs3}
casebaseAbsolute <- absoluteRisk(cb.Model,
                                 time = seq(1, max(completeData$d.time), 0.5),
                                 newdata = t(newData),
                                 s = "lambda.1se", method = "numerical")
```

We will compare this curve to a regularized version of Cox regression, and a Kaplan-Meier survival curve. Using the \pkg{glmnet} package, we can fit the Cox regression model using regularized estimation. However, \pkg{glmnet} does not provide any functionality for estimating the baseline hazard, and the functionalities in the \pkg{survival} package do not work on the output from \code{cv.glmnet}. To circumvent this issue and estimate the baseline hazard, we followed the following steps. First, we fitted a non-regularized Cox model using only the variables selected by \pkg{glmnet}; the baseline hazard will be estimated using this object. Then, we replaced the estimates of the regression coefficients by the estimates from \pkg{glmnet}. Finally, this modified \code{coxph} object is fed to the \code{survival::survfit} function to obtain the estimated survival function.

```{r coxHazAbsolute, echo = TRUE, eval = eval_cs3}
# Create u and xCox, which will be used when fitting Cox with glmnet.
u <- survival::Surv(time = as.numeric(y[,2]), 
                    event = as.factor(y[,1]))
xCox <- as.matrix(sparse.model.matrix(death~ . - d.time,
                                      data = completeData))[-c(sam),]
# hazard for cox using glmnet
coxglmFit <- glmnet::cv.glmnet(x = xCox, y = u,
                               family = "cox", alpha = 1)
# convergence demonstrated in plot
# plot(coxglmFit)
#taking the coefficient estimates for later use
nonzero_covariate_cox <- predict(coxglmFit, type = "nonzero", s = "lambda.1se")
nonzero_coef_cox <- coef(coxglmFit, s = "lambda.1se")
# creating a new dataset that only contains the covariates chosen through glmnet.
cleanCoxData <- as.data.frame(cbind(as.numeric(y[,2]),
                                    as.numeric(y[,1]),
                                    xCox[, nonzero_covariate_cox$X1]))
# fitting a cox model using regular estimation, however we will not keep it.
# this is used more as an object place holder.
coxFit <- survival::coxph(Surv(time = V1, event = V2) ~ ., 
                          data = cleanCoxData)
# The coefficients of this object will be replaced 
# with the estimates from coxglmFit.
coxFit$coefficients <- nonzero_coef_cox@x
# Fitting CI curve for cox+glmnet
newDataCox <- newData[nonzero_covariate_cox$X1 - 1]

abCoxFit <- survival::survfit(coxFit, newdata = as.data.frame(t(newDataCox)),
                              time = seq(0, max(completeData$d.time), 1),
                              type = "breslow")
```

We used the \pkg{survival} package to calculate the Kaplan-Meier estimate of the survival function.

```{r fitKM+abRisk, echo = TRUE, eval = eval_cs3}
km <- survival::Surv(time = completeData$d.time, 
                     event = completeData$death)
abKm <- survival::survfit(km ~ 1,
                          type = 'kaplan-meier')
```

Now that our three cumulative incidence functions have been calculated, we compare them all in Figure \ref{fig:abSC}. Both \pkg{glmnet} and \pkg{casebase} have a lower estimated cumulative incidence by the end of the study, when compared to Kaplan-Meier.

```{r abSupportComparison, echo = TRUE, eval = eval_cs3, fig.cap="\\label{fig:abSC} Compares regularized casebase in black, regularized cox in red and Kaplan meier in blue."}
plot(casebaseAbsolute, type='l', col = "black", lwd = 2, 
     main = "Support- CaseBase vs. Cox+glmnet vs KM Absolute Risk Curves", 
     xlab = "Survival-Time", ylab = "Cumulative Incidence", ylim = c(0,1))
lines(abCoxFit, col = "red", fun = "event", lwd = 3, conf.int = FALSE)
lines(abKm, col = "blue", fun = "event", lwd = 3, conf.int = FALSE)
legend("bottom", 
       legend = c("casebase", "Cox+glmnet", "KM curve"), 
       col = c("black", "red", "blue"), lty = c(1, 1, 1), bg = "gray90")
```

@@

The purpose of the next case study is to show how \pkg{casebase} can also be used for survival analysis problems that do not share the properties above. To this end, we will use the Stanford Heart Transplant Data [@clark1971cardiac,@crowley1977covariance]; we will use the version available in the \pkg{surivival} package.

@@

<!-- In the following table we provide a comparison between the Cox model and case-base sampling: -->
